Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-03-17 paper.tex
update

\item \textbf{Refunctionalization} is shown in Section~\ref{refunc} which sequentializes
  the order of abstract execution by higher-order functions.
  For clarity, we first present the refunctionalized AAM without caching. We then adopt a different caching algorithm to guarantee the termination of abstract interpretation.
  In this section, we also review pushdown control-flow analysis and examine how computable
    and precise call/return matching is obtained through these transformations.

@@

\section{Refunctionalization} \label{refunc}

@@

\section{Refunctionalized AAM}

Maybe insert to section refunc.

\subsection{Store Widening}

\subsection{Precise Call/Return Match}

show the precision of stack

On equivalence: since refunc interp does not have an explicit concept of $state$,
so 1) we can compare the final states of P4F against with refunc interp,
or 2) we can compare the summarized store of P4F against with summarized store of refunc interp.

\subsection{Termination}

show that number of continuation is finite.

\subsection{Soundness}

\subsection{Complexity}

@@

\textbf{Defunctionalization and Refunctionalization.}
Defunctionalization and refunctionalization build connections between abstract
machines and interpreters.

For deterministic languages, a reduction semantics is a structural operational
semantics in continuation style, where the reduction context is a
defunctionalized continuation.
Unfortunately, abstract abstract machines are nondeterministic, where
the nondeterminism is handled by a worklist.
We can turn this explicit nondeterminism into the reduction context of
AAM.

@@

CFA2 is the first solution that solves the return-flows problem~\cite{vardoulakis2010cfa2},
but CFA2 has several drawbacks: it works only on continuation-passing style programs,
and does not supports polyvariant analysis, in addition to an exponential time
complexity.

Pushdown control flow analysis (PDCFA) is another mechanism which maintains this precision 
through the use of a specialized state graph representing all possible stacks contained 
within the unbounded-stack machine. These state graphs are known as Dyke state graphs.

Similar to PDCFA, abstracting abstract control (AAC) is another strategy for maintaining stack precision without any loss. As shown by~\citeauthor{gilray2016pushdown}, AAC functions similar to PDFCA, including the fact that neither method requires any context sensitivity. AAC functions by utilizing continuations which are specific to both the source and target states of a call-site transition, which guarantees that no merging will occur during returns~\cite{gilray2016pushdown}.

\citeauthor{gilray2016pushdown} further proposed a polyvariant continuation-addresses allocator
for small-step AAM to achieve pushdown analysis \cite{gilray2016pushdown}.
This method is both simple to implement and computationally inexpensive.
Based on the AAM we presented in Section \ref{aam}, the only changes in the code required is
not only keeping track of the entry-point expression of callee, but also holding the target
environment when allocating a continuation addresses.
The following code shows the implementation:

\begin{lstlisting}
case class P4FContAddr(tgt: Expr, tgtEnv: Env) extends KAddr
def allocKont(tgtExpr: Expr, tgtEnv: Env): KAddr = P4FContAddr(tgtExpr, tgtEnv)
\end{lstlisting}

The target environment is used for successor states and extended from the closure's
environment (namely @val new_env = env_c + (v -> baddr)@).
No other pieces of code would need to be modified.
Notably, this change only causes constant-factor increase in time complexity of
the analysis if the store is widened.

\textbf{Abstracting Definitional Interpreters.}
Reynold's seminal paper \emph{Definitional interpreters for higher-order
  programming languages} \cite{reynolds1972definitional} shows that in
definitional interpreters, the defined language can inherit properties from the defining language.
With this insight, \citeauthor{darais2017abstracting} construct abstracting
definitional interpreters which automatically inherit the pushdown control-flow
property from its defining language
because the defined language simply uses the stack the
meta-language~\cite{darais2017abstracting}. \citeauthor{darais2017abstracting}'s abstracting
definitional interpreters work on direct-style programs, and is written in monadic style. One
of the advantages of monadic abstract interpreters is modular and composable, therefore
deploying different sensitivities or features is just applying a different monad.
Prior to that, \citeauthor{Sergey:2013:MAI:2491956.2491979} show a monadic
abstract interpreter for small-step semantics.

This paper is greatly inspired by~\citeauthor{darais2017abstracting}'s work.
Instead of simply showing the final form of abstract definitional interpreters,
we reveal that refunctionalization plays an important role for inheriting the
stack from the metalanguage.

The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written in both
monadic- and direct-style.
The difference is that ours works for A-Normal Form $\lambda$-calculus, instead of plain
$\lambda$-calculus, although our work can also be easily adopted to handle a plain
$\lambda$-calculus.

>>>>

\item \textbf{Refunctionalization} is shown in Section~\ref{sec:refunc} which sequentializes
  the order of abstract execution by higher-order functions.
  For clarity, we first present the refunctionalized AAM without caching. We then adopt a different caching algorithm to guarantee the termination of abstract interpretation.
  In this section, we also review pushdown control-flow analysis and examine how computable
    and precise call/return matching is obtained through these transformations.

@@

\section{Refunctionalization} \label{sec:refunc}

@@

\iffalse
\section{Refunctionalized AAM}
Maybe insert to section refunc.
\subsection{Store Widening}
\subsection{Precise Call/Return Match}
show the precision of stack
On equivalence: since refunc interp does not have an explicit concept of $state$,
so 1) we can compare the final states of P4F against with refunc interp,
or 2) we can compare the summarized store of P4F against with summarized store of refunc interp.
\subsection{Termination}
show that number of continuation is finite.
\subsection{Soundness}
\subsection{Complexity}
\fi

@@

\textbf{Defunctionalization and Refunctionalization.}
Defunctionalization and refunctionalization build connections between abstract
machines and interpreters.
\citeauthor{reynolds1972definitional} first shows refunctionalization as a program transformation technique that 
can be used to transform higher-order functions to first-order functions 
when sepcifing a programming language by interpreter \cite{reynolds1972definitional}.
Refunctionalization is the reversed transformation.

Over the years, Danvy and his collaborators applied refunctionalization and defunctionalization to 
many different abstract machines and semantics, including the CEK machines, the CLS machines, 
the SECD machines, and etc
\cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing, Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268, danvy2009towards, biernacka2009towards}.
And other applications are also found to be applicable, such as Dyck words recognizer and Dijkstra's shunting-yard algorithm \cite{Danvy:2006:RW:2171265.2171268}.

For deterministic languages, \citeauthor{Danvy:2008:DIP:1411204.1411206} shows 
a reduction semantics is a structural operational semantics in continuation style, 
where the reduction context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
Our work further shows for nondeterministic languages, there is another layer of continuations
that controls the nondeterministic choices. After identifing and explicitly exposing (section \ref{linear}) that layer
of nondeterministic continuations, the whole program can be then refunctionalized to extended continuation-passing style.

Fusion as one of the transformation in our paper is also studied by \citeauthor{Ohori:2007:LFF:1190216.1190241} \cite{Ohori:2007:LFF:1190216.1190241}.

@@

CFA2 is the first solution that solves the return-flows problem~\cite{vardoulakis2010cfa2},
but CFA2 has several drawbacks: it works only on continuation-passing style programs,
and does not supports polyvariant analysis, in addition to an exponential time
complexity.
Pushdown control flow analysis (PDCFA) is a mechanism which maintains this precision 
through the use of a Dyke state graph representing all possible stacks contained 
within the unbounded-stack machine\cite{earl2012introspective, earl2010pushdown}.
Similar to PDCFA, abstracting abstract control (AAC) is another strategy for 
maintaining stack precision \cite{johnson2015abstracting}.
AAC functions by utilizing continuations which are specific to both 
the source and target states of a call-site transition, 
which guarantees that no spurious merging will occur during returns.

\citeauthor{gilray2016pushdown} further proposed a polyvariant continuation-addresses allocator
for small-step AAM to achieve pushdown analysis \cite{gilray2016pushdown}.
This method is both simple to implement and computationally inexpensive and so called Pushdown for Free (P4F).
Based on the AAM we presented in Section \ref{aam}, the only changes in the code required is
not only keeping track of the entry-point expression of callee, but also holding the target
environment when allocating a continuation addresses.
No other pieces of code would need to be modified.
Notably, this change only causes constant-factor increase in time complexity of
the analysis if the store is widened.

\iffalse
The following code shows the implementation:
\begin{lstlisting}
case class P4FContAddr(tgt: Expr, tgtEnv: Env) extends KAddr
def allocKont(tgtExpr: Expr, tgtEnv: Env): KAddr = P4FContAddr(tgtExpr, tgtEnv)
\end{lstlisting}
The target environment is used for successor states and extended from the closure's
environment (namely @val new_env = env_c + (v -> baddr)@).
\fi

\textbf{Abstracting Definitional Interpreters.}
Reynold's seminal paper \emph{Definitional interpreters for higher-order
  programming languages} \cite{reynolds1972definitional} shows that in
definitional interpreters, the defined language can inherit properties from the defining language.
With this insight, \citeauthor{darais2017abstracting} construct abstracting
definitional interpreters which automatically inherit the pushdown control-flow
property from its defining language
because the defined language simply uses the stack the
meta-language~\cite{darais2017abstracting}. \citeauthor{darais2017abstracting}'s abstracting
definitional interpreters work on direct-style programs, and is written in monadic style. One
of the advantages of monadic abstract interpreters is modular and composable, therefore
deploying different sensitivities or features is just applying a different monad.
Prior to that, \citeauthor{Sergey:2013:MAI:2491956.2491979} show a monadic
abstract interpreter for small-step semantics \cite{Sergey:2013:MAI:2491956.2491979}.

This paper is greatly inspired by~\citeauthor{darais2017abstracting}'s work.
The difference on the surface is that ours works for A-Normal Form $\lambda$-calculus, 
instead of plain $\lambda$-calculus, although our work can also be easily adopted to handle a plain
$\lambda$-calculus.
But except of simply showing the final form of abstract definitional interpreters,
we reveal that refunctionalization plays an important role for inheriting the
stack from the metalanguage.
The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written direct-style with 
delimited controls. The correspondence of delimited controls (as well as continuations) and monads
are well-known \cite{Danvy:1990:AC:91556.91622, wadler1992essence, danvy1992representing, moggi1991notions}.
